From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-06-10 draft_v1/draft_v1.qmd
update_draft_Leoniefeedback

---
title: "From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study"
#subtitle: Using the HELIUS dataset
author: 
  - name: Kyuri Park
    affil-id: 1
  - name: Leonie K. Elsenburg
    affil-id: 2
  - name: Mary Nicolao
    affil-id: 2
  - name: Karien Stronks
    affil-id: 2
  - name: Vítor V. Vasconcelos
    affil-id: 1, 3

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling. Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between domain-specific precariousness indicators, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precarity under varying levels of financial stress reduction. Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise. In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions. More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.

@@

Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering]. This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk. However, despite growing evidence of association between precariousness and mental health, a fundamental question remains unresolved: *How do these factors causally influence one another?* Most existing studies rely on cross-sectional correlations, leaving open questions about directionality and feedback. Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate unintended effects.

This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precarity–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
By examining both aggregated measures (depression sum score and a composite precarity index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precarity co-evolve under varying levels of external stress. This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold. In others, it shows unistable behavior, responding more gradually to incremental changes.
By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precarity system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.

@@

<!-- This study aims to investigate the causal mechanisms linking precariousness and depression, and to examine how those mechanisms shape the system’s response to external intervention. Specifically, we ask: How do different configurations of the precarity–depression relationship affect the outcomes of interventions aimed at reducing external stress? -->

<!-- To address this, we combine causal discovery and mechanistic modeling. We apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, precariousness indicators, and depressive symptoms. Our analysis includes both aggregate scores (e.g., PHQ-9 total, precarity composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways. -->

<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precarity under varying levels of financial stress. This framework allows us to examine how internal system properties—particularly feedback strength and stochastic noise—influence responsiveness to intervention. In some regimes, the system exhibits bistability, resisting change unless pushed past a critical threshold. In others, it transitions gradually, reflecting unistable dynamics. Through this integrative approach, we aim to show how the internal architecture of the depression–precarity system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions. -->

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort]. It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and which provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement]. To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures. These analyses led to the identification of five distinct precariousness factors—three domain-based and two reflecting recent stressors—each composed of multiple observed variables, as outlined below. Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).

@@

After preprocessing, the HELIUS dataset includes 21,628 samples. In addition to the five identified precariousness factors, we computed an overall precarity score by aggregating their values. Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we explore the relationship between depression and precarity using both aggregated scores and their disaggregated components. See @fig-dist for the overall distributions of the variables used in the analysis.

```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%", cache=TRUE}
#| fig-cap: >
#|   Distributions of variables with density overlay.  
#|   
#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors; *Precarity* = overall precarity;
#|   *PHQsum* = PHQ-9 sum score; *anh* = anhedonia; *app* = appetite; *con* = concentration; *dep* = depressed mood; *ene* = energy; *glt* = guilty; *mot* = motor; *sui* = suicidal
#| label: fig-dist

clust_data <- scaled_data |> 
  mutate(
    P.emp = rowMeans(cbind(emp_stat, work_sit)),
    P.soc = rowMeans(cbind(soc_freq, soc_adq)),
    P.hou = rowMeans(cbind(nb_safe, nb_res, nb_rent, cul_rec)),
    S.rel = rowMeans(cbind(frd_brk12, conf12)),
    S.fin = rowMeans(cbind(fincri12, inc_dif)),
    Precarity = P.emp + P.soc + P.hou + S.rel + S.fin
  ) |>
  dplyr::select(17:34) |>
  as.data.frame()

@@

# Plot distributions with density lines
ggplot(clust_data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "Precarity", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
  labs(x = "", y = "") +
  theme_minimal()
```

@@

The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precarity relate to overall depression severity. However, it may obscure finer-grained effects. The disaggregated analysis allows for precise mapping of which symptoms are influenced by (or influence) specific precariousness conditions, but it introduces complexity due to the high dimensionality and distributional properties of the data. The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precarity across domains.

@@

<!-- Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions. This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately. This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis. -->
<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression.  -->

@@

## Modeling Dynamics between Precarity and Depression

@@

NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2). This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives. From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects.

@@

## Causal Structure Linking Precarity and Depression

### Depression as sum score
The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
As described in [@sec-analysis], dashed edges indicate ties, with the exact edge endpoint where the tie occurs is represented by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (*PHQsum*). Employment precarity (*P.emp*) and social precarity (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. Supporting this, the proportion matrix plot in panel (b) shows that *S.fin* has some probability of causally influencing either *P.emp* or *P.soc.*

@@

Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder. Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional. Meanwhile, housing precarity (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*. While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.

@@

<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precarity  (P.emp) and social precarity (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.   Additionally, P.emp and S.fin are identified as non-causes of housing precarity (P.hou). Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->

@@

<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders. Employment and social precarity may also be influenced by depression, either directly or through latent variables. Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways. -->

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Resulting graph of precarity factors and depression sum score using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 100%

@@

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of *S.fin*, which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that *S.fin* may cause changes in *app*, while its relationships with other symptoms, such as *ene*, *mot*, and *dep*, remain ambiguous.
Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it. This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precarity rather than the other way around. Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.

@@

Among depressive symptoms, *dep*, *glt*, and *slp* appear to be the most connected to precariousness factors, while *P.soc* has the most connections with symptoms, predominantly as a recipient rather than a driver of influence. Within the symptom network, *slp*, *sui*, and *anh* emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms. On the other hand, *con*, *mot*, *dep*, and *glt*, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network. This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity. On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.

@@

<!-- <!-- Centrality and Connectivity --> 
<!-- Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points. Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network. Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors. These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes. -->
<!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
<!-- The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs. As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms.  -->

@@

<!-- Some discrepancies, however, exist between the symptom-level and sum graphs. For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors. Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.  -->

<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph. -->
```{r}
#| label: fig-sym
#| layout-nrow: 2
#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 70%

@@

### Precarity as sum score

@@

```{r}
#| layout-nrow: 2
#| label: fig-presum
#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

@@

## Simulating the Dynamics Between Precarity and Depression Under Intervention

>>>>

---
title: "From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study"
#subtitle: Using the HELIUS dataset
author: 
  - name: Kyuri Park
    affil-id: 1
  - name: Leonie K. Elsenburg
    affil-id: 2
  - name: Mary Nicolao
    affil-id: 2
  - name: Karien Stronks
    affil-id: 2
  - name: Vítor V. Vasconcelos
    affil-id: 1, 3

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header: header.tex
    crossref:
      appendix-title: "App."
      appendix-delim: ":"
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
nocite: |
  @*
csl: apa.csl
abstract: |
  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling. Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between domain-specific precariousness indicators, and depressive symptoms. Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precariousness under varying levels of financial stress reduction. Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise. In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions. More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.

@@

Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering; @mckee2017living]. This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk. However, despite growing evidence of association between precariousness and mental health, a fundamental question remains unresolved: *How do these factors causally influence one another?* Most existing studies rely on cross-sectional correlations, leaving open questions about directionality and feedback. Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate unintended effects.

This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention. Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
To answer this, we combine two approaches. First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress. This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold. In others, it shows unistable behavior, responding more gradually to incremental changes.
By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precariousness system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.

@@

<!-- This study aims to investigate the causal mechanisms linking precariousness and depression, and to examine how those mechanisms shape the system’s response to external intervention. Specifically, we ask: How do different configurations of the precariousness–depression relationship affect the outcomes of interventions aimed at reducing external stress? -->

<!-- To address this, we combine causal discovery and mechanistic modeling. We apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, precariousness indicators, and depressive symptoms. Our analysis includes both aggregate scores (e.g., PHQ-9 total, precariousness composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways. -->

<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precariousness under varying levels of financial stress. This framework allows us to examine how internal system properties—particularly feedback strength and stochastic noise—influence responsiveness to intervention. In some regimes, the system exhibits bistability, resisting change unless pushed past a critical threshold. In others, it transitions gradually, reflecting unistable dynamics. Through this integrative approach, we aim to show how the internal architecture of the depression–precariousness system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions. -->

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort]. It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement]. 
To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures. These analyses led to the identification of five distinct precariousness factors—three domain-based and two reflecting recent stressors—each composed of multiple observed variables, as outlined below. Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).

@@

After preprocessing, the HELIUS dataset includes 21,628 individuals. In addition to the five identified precariousness factors, we computed an overall precariousness score by by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains. Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores. In the subsequent causal discovery analysis, we explore the relationship between depression and precariousness using both aggregated scores and their disaggregated components. See @fig-dist for the overall distributions of the variables used in the analysis.

```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%", cache=TRUE}
#| fig-cap: >
#|   Distributions of variables with density overlay.  
#|   *This shows the distributions of the variables used in the causal analysis, plotted as histograms with overlaid density estimates. The x-axis represents the value of each variable after preprocessing. The y-axis shows the corresponding estimated probability density.*
#|   ***P.emp*** = *employment precariousness*;  ***P.hou*** = *housing precariousness*;  ***P.soc*** = *social precariousness*;  ***S.fin*** = *recent financial stressors*;  ***S.rel*** = *recent relational stressors*;  ***precariousness*** = *overall precariousness*;  ***PHQsum*** = *PHQ-9 sum score*;  ***anh*** = *anhedonia*;  ***app*** = *appetite*;  ***con*** = *concentration*;  ***dep*** = *depressed mood*;  ***ene*** = *energy*;  ***glt*** = *guilty*;  ***mot*** = *motor*;  ***slp*** = *sleep*;  ***sui*** = *suicidal*.
#|   
#| label: fig-dist

clust_data <- scaled_data |> 
  mutate(
    P.emp = rowMeans(cbind(emp_stat, work_sit)),
    P.soc = rowMeans(cbind(soc_freq, soc_adq)),
    P.hou = rowMeans(cbind(nb_safe, nb_res, nb_rent, cul_rec)),
    S.rel = rowMeans(cbind(frd_brk12, conf12)),
    S.fin = rowMeans(cbind(fincri12, inc_dif)),
    precariousness = P.emp + P.soc + P.hou + S.rel + S.fin
  ) |>
  dplyr::select(17:34) |>
  as.data.frame()

@@

# Plot distributions with density lines
ggplot(clust_data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
  labs(x = "", y = "") +
  theme_minimal()
```

@@

The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precariousness relate to overall depression severity. However, it may obscure finer-grained effects. The disaggregated analysis allows for precise mapping of which symptoms are influenced by (or influence) specific precariousness conditions, but it introduces complexity due to the high dimensionality and distributional properties of the data. The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precariousness across domains.

@@

<!-- Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions. This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately. This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis. -->
<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precariousness and depression.  -->

@@

## Modeling Dynamics between precariousness and Depression

@@

<!-- NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2). This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives. From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects. -->

To explore optimal parameter configurations, we used the NSGA-II algorithm as implemented in the `nsga2R` package [@nsga2R]. This algorithm is inspired by natural selection and evolves a population of candidate solutions over multiple generations to balance competing objectives. We began with a population of 500 randomly generated parameter sets and allowed them to evolve over 300 generations. During this process, solutions were selected, combined, and mutated using standard settings: a crossover probability of 0.9 (which controls how often solutions are combined), a mutation probability of 0.25 (introducing random variation), and a tournament size of 2 (used to pick the best candidates for reproduction).
The result is a Pareto front—a set of solutions that are considered optimal in the sense that no single objective can be improved without worsening another. From this diverse set of trade-off solutions (see @fig-pareto), we selected the 30 parameter sets with the lowest overall error across objectives. These top-performing configurations were then used in the simulation phase to evaluate the potential effects of different interventions.

@@

## Causal Structure Linking Precariousness and Depression

### Depression as sum score
The sum score graphs provide a high-level summary of how precariousness factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
As described in [@sec-analysis], dashed edges indicate ties, and the edge endpoint at which the tie occurs is marked by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*). Employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role. While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint. Supporting this, the proportion matrix plot in panel (b) shows that *S.fin* has some probability of causally influencing either *P.emp* or *P.soc.*

@@

Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder. Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional. Meanwhile, housing precariousness (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*. While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.

@@

<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precariousness  (P.emp) and social precariousness (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.   Additionally, P.emp and S.fin are identified as non-causes of housing precariousness (P.hou). Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship. Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->

@@

<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders. Employment and social precariousness may also be influenced by depression, either directly or through latent variables. Lastly, housing precariousness does not directly impact depression but remains connected to employment precariousness and financial stress, likely mediated by direct or latent pathways. -->

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Resulting graph of precariousness factors and depression sum score using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 100%

@@

Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of *S.fin*, which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality. One exception is the stronger evidence suggesting that *S.fin* may cause changes in *app*, while its relationships with other symptoms, such as *ene*, *mot*, and *dep*, remain ambiguous.
Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it. This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precariousness rather than the other way around. Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precariousness is more often a consequence rather than a driver of depressive symptoms.

@@

Among depressive symptoms, *dep*, *glt*, and *slp* appear to be the most connected to precariousness factors, while *P.soc* has the most connections with symptoms, predominantly as a recipient rather than a driver of influence. Within the symptom network, *slp*, *sui*, and *anh* emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms. On the other hand, *con*, *mot*, *dep*, and *glt*, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network. This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precariousness. On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precariousness, reinforcing a self-sustaining dynamic between depression and precarious conditions.

@@

<!-- <!-- Centrality and Connectivity --> 
<!-- Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points. Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network. Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precariousness factors. These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes. -->
<!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
<!-- The causal structure involving precariousness factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs. As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precariousness (*P.soc*) are more likely to be influenced by depressive symptoms.  -->

@@

<!-- Some discrepancies, however, exist between the symptom-level and sum graphs. For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precariousness factors. Additionally, housing precariousness (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.  -->

<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph. -->
```{r}
#| label: fig-sym
#| layout-nrow: 2
#| fig-cap: "Resulting graph of precariousness factors and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 70%

@@

### Precariousness as sum score

@@

```{r}
#| layout-nrow: 2
#| label: fig-presum
#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
#| fig-subcap: 
#|   - "FCI PAG"
#|   - "Proportion matrix"
#| fig-align: center
#| out-width: 60%

@@

## Simulating the Dynamics Between Precariousness and Depression Under Intervention
